% ============================================================================
% Supplementary Information SI-4: Seed Compounds and Buffer Equilibria
%
% Drop-in subsection fragment. Parent SI must provide: amsmath/amssymb,
% booktabs, siunitx, mhchem, and the tcolorbox environments {evaluation},
% {concern}, {threshbox}. Bibliography handled at the parent level.
% (Parameter reference lives in si_parameters.tex.)
% ============================================================================

\subsection{Seed Compounds and Buffer Equilibria}
\label{si:seed}

The system is bootstrapped with nine compounds and four hardcoded reactions. The compounds set the chemical scope and provide the unbiased starting conditions for the kinetics ODE; the four reactions inject acid--base equilibria that the diffusion model cannot discover.

\paragraph{Seed compounds.}
Eight molecules are inserted as \texttt{is\_seed = true} compounds at the first worker's startup, under a Postgres advisory lock so only one worker performs the seeding:
\begin{itemize}
    \item $\mathrm{CH_2O}$ (formaldehyde) --- the primary substrate, the only species at non-trace initial concentration in the upstream formose-style runs;
    \item $\mathrm{H_2O}$ (water), $\mathrm{H_2}$ (rendered as \texttt{[HH]} since the MLFF has no electron concept and so $\mathrm{H}$ / $\mathrm{H^+}$ share that SMILES) --- universal small fragments loaded from XYZ files;
    \item $\mathrm{CO_2}$, $\mathrm{H_2CO_3}$, $\mathrm{HCO_3^-}$, $\mathrm{OH^-}$, $\mathrm{H_3O^+}$ --- the buffer species needed to support the manual equilibria.
\end{itemize}
Every seed compound enters with a real 3D geometry (either from \texttt{start.xyz} or the per-fragment \texttt{.xyz} files in \texttt{fragments/}), is canonicalised through RDKit so its SMILES matches what the discovery pipeline would produce, and is registered with a single PES-graph minimum. Larger buffer compounds ($\geq 3$ atoms: $\mathrm{H_2CO_3}$, $\mathrm{HCO_3^-}$) are enqueued for PES exploration and CREST conformer search; smaller ones ($\mathrm{OH^-}$, $\mathrm{H_3O^+}$) have no internal conformers.

Each seed receives an initial concentration of \SI{1}{mM} in the kinetics ODE (SI-\ref{si:kinetics}). The same baseline is applied to every runtime-discovered species so that newly-found compounds can participate in reactions at $t = 0$ instead of being trapped at exactly zero.

\paragraph{Manual buffer equilibria.}
Four acid--base reactions are seeded with literal forward and backward rate constants and tagged \texttt{discovery\_method = "manual\_equilibrium"}:

\begin{center}
\begin{tabular}{lll@{\hspace{2em}}rr}
\toprule
Name & Forward direction & & $k_\text{fwd}$ & $k_\text{bwd}$ \\
\midrule
Water autoionisation
    & $\mathrm{H_2O \to H^+ + OH^-}$
    && $1.8 \times 10^{-3}$ & $10^{10}$ \\
$\mathrm{CO_2}$ hydration
    & $\mathrm{CO_2 + H_2O \to H_2CO_3}$
    && $5.5 \times 10^{-3}$ & $178$ \\
$\mathrm{H_2CO_3}$ dissociation
    & $\mathrm{H_2CO_3 \to H^+ + HCO_3^-}$
    && $2.5 \times 10^{6}$  & $10^{10}$ \\
Proton solvation
    & $\mathrm{H^+ + H_2O \to H_3O^+}$
    && $10^{10}$ & $1.0$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent Units: $\si{s^{-1}}$ for unimolecular forward and \si{M^{-1}.s^{-1}} for bimolecular, by convention; the solver applies the appropriate mass-action exponents from the stoichiometry. The values are literature rate constants from the upstream reference implementation (\texttt{crn-exploration/lib/kinetic\_sampler.py}); they are not derived from Eyring and are \emph{not} re-scaled with temperature when the solver runs at a non-ambient $T$.

The acid--base channels are off-limits to the diffusion proposer for a structural reason: the MD-ET force field used as the energy oracle does not represent explicit electrons, so it has no concept of a proton versus a neutral hydrogen atom. Both render to the same SMILES \texttt{[HH]}, and the model's energy/force predictions for ionic species are systematically wrong in ways that would propagate through the IRC and the barrier filters. Manual equilibria sidestep the issue entirely: their kinetic role (maintaining a self-consistent pH, ensuring $\mathrm{CO_2/H_2CO_3/HCO_3^-}$ buffering, providing a reservoir of solvated protons) is delivered by the literature rate constants, and the diffusion proposer is never asked to discover them.

The rate constants are deliberately fast on the backward direction of the dissociation steps ($10^{10}\,\si{s^{-1}}$, the diffusion-limit cap; SI-\ref{si:kinetics}). The equilibria therefore reach their pH-determined ratios on the shortest timescale resolved by the integrator, so by the time discovered chemistry begins to evolve, the buffer is already at equilibrium and acts as a fast-relaxation reservoir rather than a coupled dynamic player.
